---
title: "Running the Metabolic Chamber Simulation and Plotting Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the Metabolic Chamber Simulation and Plotting Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This picks up where `vignette("04-endotherm-inputs")` left off. Before using
an animal parameterization in a landscape-scale run, it's worth calibrating
it against a known resting metabolic rate (RMR) in a controlled "metabolic
chamber" — four fixed scenarios (standing/curled posture x
variable/constant core temperature) run against a bundled synthetic
temperature ramp, independent of any real climate data.

`run_metabolic_chamber()` needs the real `Endo2022a.exe` executable. For
licensing reasons that executable is never distributed with this package or
committed to its repository — point `exe_path` at your own local copy. The
code chunks below are gated on `file.exists(exe_path)`: they run and embed
real output when knit on a machine that has the executable, and are shown
but skipped everywhere else (including CI and CRAN-style checks).

## 1. Reuse the animal parameterization

```{r}
library(ThermalScapeR)

endo_inputs <- get_endotherm_defaults()
endo_inputs$animal$species_label    <- "Male Bighorn"
endo_inputs$animal$body_mass        <- 60
endo_inputs$animal$mass_by_julday   <- rep(60, endo_inputs$model_settings$julnum)

exe_path <- "D:/Code/Niche Mapper/NicheMap_UpdatedExe/Endotherm Model and Input Files/Endo2022a.exe"
```

`run_metabolic_chamber()` always runs at `julnum = 12` (tied to the bundled
temperature-ramp data), so it doesn't matter if `endo_inputs` above was
built at a different `julnum` — it will override the value and message you
about it.

## 2. Run all four chamber scenarios

```{r, eval = file.exists(exe_path)}
result <- run_metabolic_chamber(endo_inputs, exe_path = exe_path)
result$log
```

`result` is a `"metchamber_result"`: an hourly output table per scenario
that succeeded (`result$hourplot`), body-part dimensions parsed from the
run (`result$dimensions`), and the animal's target RMR +/- its error margin
(`result$target_rmr`), used below as reference lines.

```{r, eval = file.exists(exe_path)}
result$dimensions
```

## 3. Plot standing vs. curled metabolic rate

```{r, eval = file.exists(exe_path), fig.width = 7, fig.height = 5}
plots <- plot(result)
```

Each chart compares predicted metabolic rate against air temperature for
standing vs. curled posture, with dashed reference lines at the target RMR
+/- its error margin (`endo_inputs$model_settings$thermoreg_trigger_tolerance`). A well-calibrated
parameterization crosses those lines near its thermoneutral zone; if it
doesn't, revisit `animal`, `fur`, or `physiology` in
`vignette("04-endotherm-inputs")` before moving on to a landscape-scale
run.

```{r, eval = file.exists(exe_path)}
names(plots)
```

## Next steps

Once a parameterization is calibrated here, `vignette("06-endotherm-landscape-scale")`
covers running it across a full tiled study area with
`run_endo_big_nichemap()`.
