---
title: "Running the Endotherm Model at Landscape Scale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running the Endotherm Model at Landscape Scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This picks up two threads at once: the calibrated animal parameterization
from `vignette("05-metabolic-chamber")`, and the tiled microclimate pipeline
output from `vignette("03-running-microclimf")`, for the same dummy
UI Arboretum site. `run_endo_big_nichemap()` runs `Endo2022a.exe` across
every valid cell of every tile, locally or as a SLURM array job;
`reconstruct_endo_raster()` then stitches the chunked output into
full-domain rasters. Everything here is illustrative (`eval = FALSE`) — it
needs a complete tiled `microclim_dir`, a DEM, reflectance data, and either
a local `Endo2022a.exe` or a cluster with Wine, none of which are available
when this package's vignettes are built.

## 1. Inputs this step needs

- `tile_map` — the fine-resolution tile-ID raster from `vignette("03-running-microclimf")`'s
  `create_tiles(..., return_tiles_rast = TRUE)` call (`tiles$tiles_rast`).
- `valid_cells_mask` — a 1/0/NA raster you build yourself (this package
  doesn't produce one): typically a study area or winter-range polygon
  intersected with a water/urban mask, so only ecologically relevant cells
  are simulated.
- `dem` — the same fine-resolution DEM used for tiling (`dem_fine`), used
  to look up each cell's elevation.
- `microclim_dir` — **the raw per-tile output directory `run_micro_big_nichemap()`
  wrote in vignette 03, step 3 (`"./Microclim_out"`), not the stitched
  `Stitched/` output from step 5.** `run_endo_big_nichemap()` reads each
  tile's `Microclim_Models/{period_label}/{hgt_lbl}/Tile_{NNN}_..._MicroclimModel_{period_label}.nc`
  file directly, and only supports `file_fmt = "nc"` — vignette 03 already
  ran `run_micro_big_nichemap()` with `file_fmt = "nc"`, so no changes are
  needed there.
- `refl_dir` — the parent directory of `compute_reflectance()`'s
  `out_dir_gref` (e.g. if `out_dir_gref = "./Data/Reflectance/Gref"`, pass
  `refl_dir = "./Data/Reflectance"`); ground absorptivity is looked up from
  the `Gref/` subfolder underneath it.
- `exe_path` — the same local `Endo2022a.exe` path used in vignette 05.
  Reminder: this executable is never committed to the package repository.
- `endo_inputs` — the calibrated bundle from vignette 05.

```{r}
library(terra)
library(ThermalScapeR)

dem_fine   <- rast("./Data/DEM/DEM_GLO30.tif")
dem_coarse <- rast("./Data/DEM/DEM_Coarse_UI_Arboretum.tif")

tiles <- create_tiles(
  coarse_dem        = dem_coarse,
  fine_dem          = dem_fine,
  dates             = as.Date(c("2023-07-01", "2023-08-01")),
  tile_dims         = c(3, 3),
  return_tiles_rast = TRUE
)

winter_range <- vect("./Data/WinterRange/UI_Arboretum_range.shp")
water_mask   <- rast("./Data/Masks/UI_Arboretum_water_mask.tif")

valid_cells_mask <- rasterize(winter_range, dem_fine, field = 1)
valid_cells_mask[water_mask == 1] <- NA

endo_inputs <- get_endotherm_defaults()
endo_inputs$animal$species_label    <- "Male Bighorn"
endo_inputs$animal$body_mass        <- 60
endo_inputs$animal$mass_by_julday   <- rep(60, endo_inputs$model_settings$julnum)

exe_path <- "D:/Code/Niche Mapper/NicheMap_UpdatedExe/Endotherm Model and Input Files/Endo2022a.exe"

dates_df <- data.frame(
  Start_Dates = as.Date("2023-07-01"),
  End_Dates   = as.Date("2023-08-01")
)
```

## 2. Run locally

```{r}
log_local <- run_endo_big_nichemap(
  tile_map         = tiles$tiles_rast,
  valid_cells_mask = valid_cells_mask,
  dates            = dates_df,
  microclim_dir    = "./Microclim_out",
  dem              = dem_fine,
  refl_dir         = "./Data/Reflectance",
  exe_path         = exe_path,
  output_dir       = "./Endo_out",
  wineprefix       = NULL,
  endo_inputs      = endo_inputs,
  study_area       = "UI_Arboretum",
  parallel         = TRUE,
  ncores           = 4
)
```

A "task" here is one `(tile, period)` combination — one animal per valid
cell, simulated in `chunk_size`-day blocks (default 52) across the whole
period. `wineprefix` is ignored on Windows; on Linux/macOS it needs a
prefix already initialized via `init_wine_prefix()`, or `NULL` to let each
scenario get its own private prefix. With no `clust_array_arg`/
`clust_array_size` supplied, every `(tile, period)` combination for the
dummy site's single tile runs in this one session, the same round-robin
distribution `run_micro_big_nichemap()` uses in vignette 03.

Output lands under `output_dir/[study_area/]{period_label}/`: a
`Tile_{NNN}_manifest.csv` and `Tile_{NNN}_log.csv` per tile, and each
simulated cell's chunked hourly output at
`Tile_{NNN}/Cell_{CCCCCC}/HOURPLOT_chunk{k}_{start}_{end}.csv` —
`reconstruct_endo_raster()` (step 4 below) discovers and assembles these
automatically.

## 3. Run on a SLURM cluster

Same hidden-`...`-argument pattern as `run_micro_big_nichemap()`:
`clust_array_arg`/`clust_array_size` are passed in by an SBATCH array
script, never by `run_endo_big_nichemap()` submitting jobs itself.

`run_endo_landscape_cluster.R`:
```r
#!/usr/bin/env Rscript
library(terra)
library(ThermalScapeR)

args <- commandArgs(trailingOnly = TRUE)
get_arg <- function(name, default = NA) {
  x <- grep(paste0("^--", name, "="), args, value = TRUE)
  if (length(x) == 0) return(default)
  sub(paste0("^--", name, "="), "", x)
}
clust_array_arg  <- get_arg("clust_array_arg")
clust_array_size <- get_arg("clust_array_size")
if (is.na(clust_array_arg) || is.na(clust_array_size)) {
  stop("Both --clust_array_arg and --clust_array_size are required (missing or ",
       "misspelled flag?). Got clust_array_arg=", clust_array_arg,
       ", clust_array_size=", clust_array_size)
}

dem_fine   <- rast("./Data/DEM/DEM_GLO30.tif")
dem_coarse <- rast("./Data/DEM/DEM_Coarse_UI_Arboretum.tif")

tiles <- create_tiles(
  coarse_dem = dem_coarse, fine_dem = dem_fine,
  dates      = as.Date(c("2023-07-01", "2023-08-01")),
  tile_dims  = c(3, 3), return_tiles_rast = TRUE
)

winter_range <- vect("./Data/WinterRange/UI_Arboretum_range.shp")
water_mask   <- rast("./Data/Masks/UI_Arboretum_water_mask.tif")
valid_cells_mask <- rasterize(winter_range, dem_fine, field = 1)
valid_cells_mask[water_mask == 1] <- NA

endo_inputs <- get_endotherm_defaults()
endo_inputs$animal$species_label    <- "Male Bighorn"
endo_inputs$animal$body_mass        <- 60
endo_inputs$animal$mass_by_julday   <- rep(60, endo_inputs$model_settings$julnum)

dates_df <- data.frame(
  Start_Dates = as.Date("2023-07-01"),
  End_Dates   = as.Date("2023-08-01")
)

# Never committed to the repo -- point this at wherever the cluster node
# stores its own copy of the executable.
exe_path <- Sys.getenv("ENDO_EXE_PATH")

log <- run_endo_big_nichemap(
  tile_map = tiles$tiles_rast, valid_cells_mask = valid_cells_mask,
  dates = dates_df, microclim_dir = "./Microclim_out", dem = dem_fine,
  refl_dir = "./Data/Reflectance", exe_path = exe_path,
  output_dir = "./Endo_out", wineprefix = "~/wine_endo",
  endo_inputs = endo_inputs, study_area = "UI_Arboretum",
  headless = TRUE,
  clust_array_arg  = as.integer(clust_array_arg),
  clust_array_size = as.integer(clust_array_size)
)

cat(sprintf("Task %s/%s complete.\n", clust_array_arg, clust_array_size))
```

`run_endo_landscape.sbatch`:
```bash
#!/bin/bash
#SBATCH --job-name=endo_UI_Arboretum
#SBATCH --array=1-4
#SBATCH --cpus-per-task=1
#SBATCH --mem=8G
#SBATCH --time=04:00:00
#SBATCH --output=logs/endo_%A_%a.out

module load R

Rscript run_endo_landscape_cluster.R \
  --clust_array_arg=${SLURM_ARRAY_TASK_ID} \
  --clust_array_size=4
```

`run_endo_big_nichemap()` enumerates all `(tile, period)` combinations and
distributes them round-robin across the array, exactly like
`run_micro_big_nichemap()` in vignette 03 — size `--array` and
`--clust_array_size` to `N_tiles x N_periods`, or fewer if running multiple
combinations per node. `headless = TRUE` wraps each Wine invocation in
`xvfb-run`, needed on compute nodes with no display server; on Windows
nodes this and `wineprefix` are ignored entirely.

## 4. Reconstruct landscape-scale rasters

```{r}
recon_log <- reconstruct_endo_raster(
  root_dir   = "./Endo_out/UI_Arboretum",
  tile_map   = tiles$tiles_rast,
  dates      = dates_df,
  variable   = "metabolic_rate",
  output_dir = "./Endo_out/Stitched",
  study_area = "UI_Arboretum",
  parallel   = TRUE,
  ncores     = 4
)
recon_log
```

`variable` selects one column from each cell's trimmed hourly output —
`"metabolic_rate"` for `MET(W)`, `"water_loss"` for `EVP(G/S)`; call again
with `variable = "water_loss"` for the second raster. Each call writes one
NetCDF file, named `{study_area}_{variable}_{period_label}.nc` (e.g.
`UI_Arboretum_metabolic_rate_20230701_to_20230801.nc`), with every valid
cell's full hourly time series placed at its tile-map location — a cell
with no covering data for the period is left entirely `NA`, not treated as
an error.

## Next steps

That completes the endotherm arc across all three vignettes:
`04-endotherm-inputs` built `endo.dat`/`alomvars.dat`/`JULDAYS.dat` for a
single animal parameterization, `05-metabolic-chamber` calibrated it
against a known resting metabolic rate, and this vignette ran it across
every valid cell of a tiled study area and reconstructed full-domain
metabolic-rate and water-loss rasters ready for downstream habitat or
energetics analysis.
